Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-12-05 paper/main.tex
added Misha to author and did spell checker

\abstract
%  menntion that only one galactic model should be sufficient to avoid the criticism
  {Thin stellar streams, such as those resulting from the tidal disruption of globular clusters, have long been known and used as probes of the gravitational potential of our Galaxy, both its visible and dark contents. In particular, the presence of under-density regions, or gaps, along these streams is commonly interpreted as being due to the close passage of dark matter sub-haloes. }
 % aims heading (mandatory)
  {In this work, we investigate the perturbations induced on streams by the passage of dense stellar systems, such as globular clusters themselves, to test the possibility that they too may cause the formation of gaps. In particular, we focus on the study of the stream of Palomar 5, a well-known globular cluster in the Galactic halo, which has particularly long tidal tails.  }
 % methods heading (mandatory)
  {For this purpose, we used a particle-test code to simulate the dynamic evolution of Palomar~5 and its tidal tails when Pal~5 is subject to the gravitational action of the Galaxy and its whole system of globular clusters.}
 % results heading (mandatory)
  {Our study shows that the tails of Pal~5 are strongly perturbed by the close passage of other clusters, in particular of NGC~2808, NGC~7078, NGC~104, and that these perturbations induce the formation of gaps in the tails.}
 % conclusions heading (optional), leave it empty if necessary
  {This result suggests that baryonic dense stellar systems such as globular clusters are capable of causing gaps in streams, as already demonstrated in the case of galaxy bar-induced perturbations and giant molecular clouds, and reinforce the idea that the effect of these components should be fully addressed when interpreting the morphology and density structures of observed stellar streams.}

@@

  Stellar streams are several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occured well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters. Interestingly, this work was nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way.

@@

  \citet{2024arXiv240519410B} provides a review of stellar stream astronomy, which has enetered a new era since the publication of the data from the Gaia astrometric mission \citep{2016A&A...595A...1G}. Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and, for the brightest stars, spectroscopy. The possibility given by Gaia to track stars with coherent movements, over the entirety of the sky, has led to the discovery of dozens of new streams. For example, \citet{2018MNRAS.477.4063M} developed the \texttt{streamfinder} algorithm and applied it across a series works of \citep{2018MNRAS.481.3442M,  2018ApJ...865...85I, 2019ApJ...872..152I} to discover the series of streams: Slidr, Sylgr, Ylgr, Fimbulthul, Svol, Gjoll, Leiptr, Phlegeton and Fjorm.  According to the most recent estimates provided by \citet{2020MNRAS.493.4978S} Fjorm in particular has an extension of more than 150 degrees on the sky which has a projected length more than five times greater than that of Pal~5.

  The possibility to couple Gaia data with spectroscopic surveys, has extended the study of stellar streams beyond the quantification of their orbital properties, to a full chemical characterization. For example \citet{2019MNRAS.490.3508L} presents $S^5$, and is one example of a cross-catalogue among other works \citep{2020AJ....160..181J, 2021ApJ...911..149L, 2022ApJ...928...30L, 2024MNRAS.529.2413U}. Currently, there are about a hundred stellar streams are currently known in our Galaxy. \citet{2023MNRAS.520.5225M} provides a compilation of their tracks on the sky. Interestingly, only about 20 streams associated to known Galactic globular clusters.

  One of the interests in studying stellar streams is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. Compared to the measurement of the HI rotation curve, stellar streams offer the opportunity to investigate the potential of the host galaxy over a wide range of distances, reaching the outermost regions of the halo. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of dark matter haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams yield the most information about the galactic potential. However, the use of single streams for constraining the potential led to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, 2005ApJ...619..800J, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the distances of the Palomar~5 and GD~1 streams. Probably only by using the set of known streams can the degeneracy of the solution be eliminated. This approach has been recently used by \citet{2024ApJ...967...89I} who constructed the most accurate model of the Milky Way.

@@

  Beyond the global (visible and dark) mass distribution, as we anticipated earlier in this Introduction, streams can be used also to infer the granularity of the dark matter, that is the mass and density of the sub-halos populating our Galaxy. According to simulations by \citet{2008MNRAS.391.1685S}, $\Lambda$-CDM predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A} ]{2005JCAP...08..003G}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars.  \citet{2012ApJ...760...75C, 2013ApJ...768..171C} applied this idea and looked for the presence of gaps in the known Palomar~5 and GD~1 streams, concluding that the density variations found in their streams were consistent with expectations from $\Lambda$~CDM models.  \citet{2019ApJ...880...38B} provided further observational evidence for this idea, identifying a gap and a spur in the GD-1 stream that they could not explain by known objects, such as globular clusters, and which they suggested was due to the close passage of a dark matter sub-halo. Interestingly enough, as stated in \citet{2019ApJ...880...38B}, the recovered properties of this sub-halo (mass and size) were inconsistent with the $\Lambda$~CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}. The above-mentioned works are only a few of a much more extensive literature that has explored the impact of dark matter sub-haloes in simulated streams \citep{2016ApJ...828L..10H, 2021MNRAS.507.1999H, 2021JCAP...10..043B, 2024arXiv240402953H, 2024arXiv241021174N}, or searched for their traces in observed streams \citep{2016MNRAS.460.2711T, 2017MNRAS.470...60E, 2020ApJ...889...70B, 2020ApJ...892L..37B}.

  In contrast, a more limited number of works have been interested in understanding whether other structures, e.g. from baryonic matter, can cause variations in the density of streams and gaps that can be assimilated and confused with those produced by dark matter sub-haloes.  Among these works, it is worth mentioning the results of \citet{2017NatAs...1..633P} that suggests how the presence of the bar at the center of the Galaxy can perturb the characteristics of a stream such as Pal~5 and generate gaps along its tail. That the Galactic bar could have an influence on streams morphology was also discussed by \citet{2016MNRAS.460..497H} and \citet{2016ApJ...824..104P}, in the case of the Ophiuchus stream \citep{2014MNRAS.443L..84B}. Besides the Galactic bar, also giant molecular clouds can produce gaps in stellar streams, as shown by \citet{2016MNRAS.463L..17A}. All these works thus seem to indicate that baryonic structures can play an important role in tail morphology. In this context,  an extensive numerical study, specifically focused on modelling the tails of Palomar~5, under the influence of the Galactic bar, spiral arms, giant molecular clouds and globular clusters, has been realised by \citet{2019MNRAS.484.2009B}, who concluded that both the influence of the bar and that of the giant molecular clouds can leave imprints in Palomar~5 tidal tails similar to those left by dark matter sub-halos, while the effect of globular clusters was found to be negligible.

  Few studies have specifically investigated the effect of globular clusters on stellar streams. The work of \citet{2017MNRAS.470...60E} concluded that the observed density variations found in Palomar~5 tails could not be due to the passage of globular clusters. However, note that this conclusion was not based on a numerical modelling of this effect. More recently \citet{2022ApJ...941..129D} have examined the possibility that the gaps found in the GD-1 stellar stream could be due to the close passage of globular clusters, concluding that this scenario is highly unlikely. These first works seem to suggest that the impact of globular clusters on stellar streams is negligible, and this is somewhat surprising, especially for streams of clusters as Palomar~5, which are confined in the inner 20~kpc of the Galaxy, where many other globular clusters do also orbit. Additionally, \citet{2023A&A...678A..69I} showed that globular cluster can even collide with other clusters, which implies that cluster stream collisions should happen much more frequently since streams are far more extended than clusters.

  In the present study we wish to fill this gap in the literature of numerically modeling cluster-stream interactions. We seak to quantify the importance of the effects of passing globular clusters in the vicinity of streams, to understand whether these systems can also be effective - and how often - in altering the distribution of stars in the tails and producing under-dense regions or gaps. To this end, on the following pages we will present the results of simulations of the streams of Palomar~5, subject to the gravitational interaction with the set of 165 Galactic globular clusters for which positions and velocities are known to date, and for which orbits can therefore be reconstructed. We chose to simulate Palomar~5 because it is a typical halo cluster with extended tails, and because it is a cluster for which the effect of baryonic structures on its stream has already been studied. As we will show, and in tension with previous claims, the close passage of other clusters with the stream of Pal~5 is not a rare event, and indeed, in the 50 simulations we ran, we find the formation of numerous gaps, for an average of 1.5 gaps per simulation, generated by 18 different clusters among the whole system of Galactic globular clusters.

  \begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{stream_on_sky_Pal5_monte-carlo-009_pouliasis2017pii-GCNBody_pouliasis2017pii.png}
    \caption{Simulated Palomar 5 sstream created by modeling the host cluster as a Plummer sphere disrupting within an axis-symmetric galactic potential plus the gravitational effect of 164 other galactic globular clusters. The top panel shows the distribution of star-particles that escaped the cluster due to tidal forces. The bottom panel shows the 1D density profile marginalized over longitude. The gray fill shows a reference simulation where the same conditions to produce the stream are used but the effect of other globular clusters are not included. The large central peak in density are of the particles still bound to Palomar~5. $\ell_0,b_0$ are located Palomar~5's cluster's center of mass. Two large gaps are present and are due to the passage of two globular clusters. $N_p$ indicates the number of particles in a bin while $N$ is the total number of particles which is 100,000.}
    \label{fig:stream_on_sky}
    \end{figure*}

@@

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in \citet{2023A&A...673A..44F}, which we present here for completeness. We begin by extracting positions in the sky, proper motions, line-of-sight velocities and distances, as well as masses and half-mass radii, of 165 globular clusters from the Galactic globular cluster catalog by \cite{2021MNRAS.505.5957B}.\footnote{The Baumgardt catalog has been assembled across a series of works, see: \cite{2020PASA...37...46B,2019MNRAS.482.5138B,2018MNRAS.478.1520B}. The catalog can be found on the World Wide Web at \href{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}.} We then convert the initial conditions from sky coordinates into a Galactocentric reference frame, by adopting a velocity for the local standard of rest of $v_{\text{LSR}} = 240$~km~s$^{-1}$ and a peculiary velocity of the Sun equal to $(U_\odot, V_\odot, W_\odot)=(11.1, 12.24, 7.25)$~km~s$^{-1}$, as reported by \citet{2012MNRAS.427..274S}.  The position of the Sun was set to $(x_\odot,y_\odot,z_\odot) = (-8.34,0,0.027)$~kpc, vertical position above the disk was taken from \citet{2001ApJ...553..184C} and the distance of the Sun to the Galactic center was taken from \citet{2014ApJ...783..130R}. These transformations were performed using \texttt{astropy} \citep{2013A&A...558A..33A}.

@@

    \subsubsection*{ \texttt{Full} simulations}
      There is a primary methodological departure from \citet{2023A&A...673A..44F}.  In that paper, globular clusters evolved under the gravitational effect of the Galaxy only, while now we include also the effect of all the other Galactic globular clusters, by taking into account the direct N-body interactions among them. To quantify these interactions, all clusters are represented by Plummer spheres, each with mass and half-mass radius as reported in the Baumgardt catalogue \citep{2021MNRAS.505.5957B}. In the following of this paper, we refer to these simulations, where the globular clusters interactions are taken into account as \texttt{full} simulations. \\
    For these simulations, we proceed into two steps:
      \begin{enumerate}
        \item First, starting from the Galactocentric positions and velocities of all 165 Galactic globular clusters, we integrate their orbits back in time for 5~Gyr under the influence of the Galaxy itself, and their mutual influence. In the backward integration, the equations of motion for the globular cluster system are thus: 
        \begin{equation}
          \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
        \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galactocentric position vector, the index $i$ indicates the globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates; the index $j$ indicates the other globular clusters that are summed over. $N_{GC}$ is the total number of globular clusters which in this study is 165, $m_j$ is the mass of the j-th cluster in the sample, $b_j$ is its Plummer scale radius, and $\vec{r_j}$ is its galactocentric position. $\Phi$ represents the Galactic smooth potential \citep[][Model~II, in the present case]{2017A&A...598A..66P}. Note that the masses and sizes of the globular clusters are kept constant in these simulations and not allowed to vary with time. This implies that we do not take into account in these models the internal evolution of the globular clusters, nor their response to the Galactic tidal field.

        \item Once the positions and velocities of the entire globular cluster system, as to 5~Gyr ago, are found, we sample Palomar~5 with  100,000 particles which follow a Plummer distribution of mass and half-mass radius as in the Baumgardt catalogue.  We then integrate the evolution of these particles forward in time to the present day, by taking into account that each particle feels the gravitational potential of the Galaxy, that of its host cluster, and that of all the other clusters in the Galaxy. Note that we do not take into account self-gravity among particles. The particles expereince the gravitational field yet do not contribute to it. The equation of motion of a generic particle among the 100,000 which populate Pal~5 is thus: 
        \begin{equation}
          \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
        \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, $\vec{r_p}$ being its position, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}. Note in that Eq.~\ref{eq:equation_of_motion_particle} the positions of the globular clusters are time-dependent since they are being loaded during this step and not computed, unlike Eq.~\ref{eq:equation_of_motion_particle}.

@@

      The procedure described so far has been repeated 50 times, each time generating a new set of initial conditions, given the uncertainties on proper motions, line-of-sight velocities, distances to the Sun, and masses, of all clusters, as reported in the Baumgardt catalogue. We handle these uncertainties  through a Monte-carlo approach, by sampling them with gaussian distribution, and by taking into account the covariance term between the proper motions. For the first simulation, we use the most probable values for the initial conditions.

      During the integration, intermediate snapshots were saved to facilitate the analysis of stellar streams and the effects of cluster impacts. Specifically, for each of the 50 realizations of the Palomar 5 stream, we saved $5000$ intermediate timesteps which is a temporal resolution of 1 million years. We provide the parameters that specify our data volumne in Table~\ref{tab:data_volume}. Using single precision floating point numbers, the size of our simulations is approximately:
    \begin{equation} \label{eq:data_volume_estimate}
      N_p \times N_{\textrm{ts}}\times N_{\textrm{phase}}\times N_{\textrm{sampling}} \times 4~\textrm{bytes}\approx 600~\textrm{Gb}.
    \end{equation}

>>>>

\abstract
%  menntion that only one galactic model should be sufficient to avoid the criticism
  {Thin stellar streams, such as those resulting from the tidal disruption of globular clusters, have long been known and used as probes of the gravitational potential of our Galaxy, both its visible and dark contents. In particular, the presence of under-density regions, or gaps, along these streams is commonly interpreted as being due to the close passage of dark matter sub-halos. }
 % aims heading (mandatory)
  {In this work, we investigate the perturbations induced on streams by the passage of dense stellar systems, such as globular clusters themselves, to test the possibility that they too may cause the formation of gaps. In particular, we focus on the study of the stream of Palomar 5, a well-known globular cluster in the Galactic halo, which has particularly long tidal tails.  }
 % methods heading (mandatory)
  {For this purpose, we used a particle-test code to simulate the dynamic evolution of Palomar~5 and its tidal tails when Pal~5 is subject to the gravitational action of the Galaxy and its whole system of globular clusters.}
 % results heading (mandatory)
  {Our study shows that the tails of Pal~5 are strongly perturbed by the close passage of other clusters, in particular of NGC~2808, NGC~7078, NGC~104, and that these perturbations induce the formation of gaps in the tails.}
 % conclusions heading (optional), leave it empty if necessary
  {This result suggests that baryonic dense stellar systems such as globular clusters are capable of causing gaps in streams, as already demonstrated in the case of galaxy bar-induced perturbations and giant molecular clouds, and reinforce the idea that the effect of these components should be fully addressed when interpreting the morphology and density structures of observed stellar streams.}

@@

  Stellar streams are several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occurred well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters. Interestingly, this work was nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way.

@@

  \citet{2024arXiv240519410B} provides a review of stellar stream astronomy, which has entered a new era since the publication of the data from the Gaia astrometric mission \citep{2016A&A...595A...1G}. Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and, for the brightest stars, spectroscopy. The possibility given by Gaia to track stars with coherent movements, over the entirety of the sky, has led to the discovery of dozens of new streams. For example, \citet{2018MNRAS.477.4063M} developed the \texttt{streamfinder} algorithm and applied it across a series works of \citep{2018MNRAS.481.3442M,  2018ApJ...865...85I, 2019ApJ...872..152I} to discover the series of streams: Slidr, Sylgr, Ylgr, Fimbulthul, Svol, Gjoll, Leiptr, Phlegeton and Fjorm.  According to the most recent estimates provided by \citet{2020MNRAS.493.4978S} Fjorm in particular has an extension of more than 150 degrees on the sky which has a projected length more than five times greater than that of Pal~5.

  The possibility to couple Gaia data with spectroscopic surveys, has extended the study of stellar streams beyond the quantification of their orbital properties, to a full chemical characterization. For example \citet{2019MNRAS.490.3508L} presents $S^5$, and is one example of a cross-catalog among other works \citep{2020AJ....160..181J, 2021ApJ...911..149L, 2022ApJ...928...30L, 2024MNRAS.529.2413U}. Currently, there are about a hundred stellar streams are currently known in our Galaxy. \citet{2023MNRAS.520.5225M} provides a compilation of their tracks on the sky. Interestingly, only about 20 streams associated to known Galactic globular clusters.

  One of the interests in studying stellar streams is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. Compared to the measurement of the HI rotation curve, stellar streams offer the opportunity to investigate the potential of the host galaxy over a wide range of distances, reaching the outermost regions of the halo. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of dark matter halos using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams yield the most information about the galactic potential. However, the use of single streams for constraining the potential led to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, 2005ApJ...619..800J, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the distances of the Palomar~5 and GD~1 streams. Probably only by using the set of known streams can the degeneracy of the solution be eliminated. This approach has been recently used by \citet{2024ApJ...967...89I} who constructed the most accurate model of the Milky Way.

@@

  Beyond the global (visible and dark) mass distribution, as we anticipated earlier in this Introduction, streams can be used also to infer the granularity of the dark matter, that is the mass and density of the sub-halos populating our Galaxy. According to simulations by \citet{2008MNRAS.391.1685S}, $\Lambda$-CDM predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{sub-halos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A} ]{2005JCAP...08..003G}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter sub-halos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that sub-halos could create gaps in stellar streams during flyby encounters, where a sub-halo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars.  \citet{2012ApJ...760...75C, 2013ApJ...768..171C} applied this idea and looked for the presence of gaps in the known Palomar~5 and GD~1 streams, concluding that the density variations found in their streams were consistent with expectations from $\Lambda$~CDM models.  \citet{2019ApJ...880...38B} provided further observational evidence for this idea, identifying a gap and a spur in the GD-1 stream that they could not explain by known objects, such as globular clusters, and which they suggested was due to the close passage of a dark matter sub-halo. Interestingly enough, as stated in \citet{2019ApJ...880...38B}, the recovered properties of this sub-halo (mass and size) were inconsistent with the $\Lambda$~CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}. The above-mentioned works are only a few of a much more extensive literature that has explored the impact of dark matter sub-halos in simulated streams \citep{2016ApJ...828L..10H, 2021MNRAS.507.1999H, 2021JCAP...10..043B, 2024arXiv240402953H, 2024arXiv241021174N}, or searched for their traces in observed streams \citep{2016MNRAS.460.2711T, 2017MNRAS.470...60E, 2020ApJ...889...70B, 2020ApJ...892L..37B}.

  In contrast, a more limited number of works have been interested in understanding whether other structures, e.g. from baryonic matter, can cause variations in the density of streams and gaps that can be assimilated and confused with those produced by dark matter sub-halos.  Among these works, it is worth mentioning the results of \citet{2017NatAs...1..633P} that suggests how the presence of the bar at the center of the Galaxy can perturb the characteristics of a stream such as Pal~5 and generate gaps along its tail. That the Galactic bar could have an influence on streams morphology was also discussed by \citet{2016MNRAS.460..497H} and \citet{2016ApJ...824..104P}, in the case of the Ophiuchus stream \citep{2014MNRAS.443L..84B}. Besides the Galactic bar, also giant molecular clouds can produce gaps in stellar streams, as shown by \citet{2016MNRAS.463L..17A}. All these works thus seem to indicate that baryonic structures can play an important role in tail morphology. In this context,  an extensive numerical study, specifically focused on modeling the tails of Palomar~5, under the influence of the Galactic bar, spiral arms, giant molecular clouds and globular clusters, has been realized by \citet{2019MNRAS.484.2009B}, who concluded that both the influence of the bar and that of the giant molecular clouds can leave imprints in Palomar~5 tidal tails similar to those left by dark matter sub-halos, while the effect of globular clusters was found to be negligible.

  Few studies have specifically investigated the effect of globular clusters on stellar streams. The work of \citet{2017MNRAS.470...60E} concluded that the observed density variations found in Palomar~5 tails could not be due to the passage of globular clusters. However, note that this conclusion was not based on a numerical modeling of this effect. More recently \citet{2022ApJ...941..129D} have examined the possibility that the gaps found in the GD-1 stellar stream could be due to the close passage of globular clusters, concluding that this scenario is highly unlikely. These first works seem to suggest that the impact of globular clusters on stellar streams is negligible, and this is somewhat surprising, especially for streams of clusters as Palomar~5, which are confined in the inner 20~kpc of the Galaxy, where many other globular clusters do also orbit. Additionally, \citet{2023A&A...678A..69I} showed that globular cluster can even collide with other clusters, which implies that cluster stream collisions should happen much more frequently since streams are far more extended than clusters.

  In the present study we wish to fill this gap in the literature of numerically modeling cluster-stream interactions. We seek to quantify the importance of the effects of passing globular clusters in the vicinity of streams, to understand whether these systems can also be effective - and how often - in altering the distribution of stars in the tails and producing under-dense regions or gaps. To this end, on the following pages we will present the results of simulations of the streams of Palomar~5, subject to the gravitational interaction with the set of 165 Galactic globular clusters for which positions and velocities are known to date, and for which orbits can therefore be reconstructed. We chose to simulate Palomar~5 because it is a typical halo cluster with extended tails, and because it is a cluster for which the effect of baryonic structures on its stream has already been studied. As we will show, and in tension with previous claims, the close passage of other clusters with the stream of Pal~5 is not a rare event, and indeed, in the 50 simulations we ran, we find the formation of numerous gaps, for an average of 1.5 gaps per simulation, generated by 18 different clusters among the whole system of Galactic globular clusters.

  \begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{stream_on_sky_Pal5_monte-carlo-009_pouliasis2017pii-GCNBody_pouliasis2017pii.png}
    \caption{Simulated Palomar 5 stream created by modeling the host cluster as a Plummer sphere disrupting within an axis-symmetric galactic potential plus the gravitational effect of 164 other galactic globular clusters. The top panel shows the distribution of star-particles that escaped the cluster due to tidal forces. The bottom panel shows the 1D density profile marginalized over longitude. The gray fill shows a reference simulation where the same conditions to produce the stream are used but the effect of other globular clusters are not included. The large central peak in density are of the particles still bound to Palomar~5. $\ell_0,b_0$ are located Palomar~5's cluster's center of mass. Two large gaps are present and are due to the passage of two globular clusters. $N_p$ indicates the number of particles in a bin while $N$ is the total number of particles which is 100,000.}
    \label{fig:stream_on_sky}
    \end{figure*}

@@

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in \citet{2023A&A...673A..44F}, which we present here for completeness. We begin by extracting positions in the sky, proper motions, line-of-sight velocities and distances, as well as masses and half-mass radii, of 165 globular clusters from the Galactic globular cluster catalog by \cite{2021MNRAS.505.5957B}.\footnote{The Baumgardt catalog has been assembled across a series of works, see: \cite{2020PASA...37...46B,2019MNRAS.482.5138B,2018MNRAS.478.1520B}. The catalog can be found on the World Wide Web at \href{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}.} We then convert the initial conditions from sky coordinates into a Galactocentric reference frame, by adopting a velocity for the local standard of rest of $v_{\text{LSR}} = 240$~km~s$^{-1}$ and a peculiar velocity of the Sun equal to $(U_\odot, V_\odot, W_\odot)=(11.1, 12.24, 7.25)$~km~s$^{-1}$, as reported by \citet{2012MNRAS.427..274S}.  The position of the Sun was set to $(x_\odot,y_\odot,z_\odot) = (-8.34,0,0.027)$~kpc, vertical position above the disk was taken from \citet{2001ApJ...553..184C} and the distance of the Sun to the Galactic center was taken from \citet{2014ApJ...783..130R}. These transformations were performed using \texttt{astropy} \citep{2013A&A...558A..33A}.

@@

    \subsubsection*{ \texttt{Full} simulations}
      There is a primary methodological departure from \citet{2023A&A...673A..44F}.  In that paper, globular clusters evolved under the gravitational effect of the Galaxy only, while now we include also the effect of all the other Galactic globular clusters, by taking into account the direct N-body interactions among them. To quantify these interactions, all clusters are represented by Plummer spheres, each with mass and half-mass radius as reported in the Baumgardt catalog \citep{2021MNRAS.505.5957B}. In the following of this paper, we refer to these simulations, where the globular clusters interactions are taken into account as \texttt{full} simulations. \\
    For these simulations, we proceed into two steps:
      \begin{enumerate}
        \item First, starting from the Galactocentric positions and velocities of all 165 Galactic globular clusters, we integrate their orbits back in time for 5~Gyr under the influence of the Galaxy itself, and their mutual influence. In the backward integration, the equations of motion for the globular cluster system are thus: 
        \begin{equation}
          \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
        \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galactocentric position vector, the index $i$ indicates the globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates; the index $j$ indicates the other globular clusters that are summed over. $N_{GC}$ is the total number of globular clusters which in this study is 165, $m_j$ is the mass of the j-th cluster in the sample, $b_j$ is its Plummer scale radius, and $\vec{r_j}$ is its galactocentric position. $\Phi$ represents the Galactic smooth potential \citep[][Model~II, in the present case]{2017A&A...598A..66P}. Note that the masses and sizes of the globular clusters are kept constant in these simulations and not allowed to vary with time. This implies that we do not take into account in these models the internal evolution of the globular clusters, nor their response to the Galactic tidal field.

        \item Once the positions and velocities of the entire globular cluster system, as to 5~Gyr ago, are found, we sample Palomar~5 with  100,000 particles which follow a Plummer distribution of mass and half-mass radius as in the Baumgardt catalog.  We then integrate the evolution of these particles forward in time to the present day, by taking into account that each particle feels the gravitational potential of the Galaxy, that of its host cluster, and that of all the other clusters in the Galaxy. Note that we do not take into account self-gravity among particles. The particles experience the gravitational field yet do not contribute to it. The equation of motion of a generic particle among the 100,000 which populate Pal~5 is thus: 
        \begin{equation}
          \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
        \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, $\vec{r_p}$ being its position, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}. Note in that Eq.~\ref{eq:equation_of_motion_particle} the positions of the globular clusters are time-dependent since they are being loaded during this step and not computed, unlike Eq.~\ref{eq:equation_of_motion_particle}.

@@

      The procedure described so far has been repeated 50 times, each time generating a new set of initial conditions, given the uncertainties on proper motions, line-of-sight velocities, distances to the Sun, and masses, of all clusters, as reported in the Baumgardt catalog. We handle these uncertainties  through a Monte-Carlo approach, by sampling them with Gaussian distribution, and by taking into account the covariance term between the proper motions. For the first simulation, we use the most probable values for the initial conditions.

      During the integration, intermediate snapshots were saved to facilitate the analysis of stellar streams and the effects of cluster impacts. Specifically, for each of the 50 realizations of the Palomar 5 stream, we saved $5000$ intermediate time-steps which is a temporal resolution of 1 million years. We provide the parameters that specify our data volume in Table~\ref{tab:data_volume}. Using single precision floating point numbers, the size of our simulations is approximately:
    \begin{equation} \label{eq:data_volume_estimate}
      N_p \times N_{\textrm{ts}}\times N_{\textrm{phase}}\times N_{\textrm{sampling}} \times 4~\textrm{bytes}\approx 600~\textrm{Gb}.
    \end{equation}

    \begin{table}[h]
      \centering
      \begin{tabular}{|c|c|c|c|}
          \hline
          $N_p$ & $N_{\textrm{ts}}$ & $N_{\textrm{phase}}$ & $N_{\textrm{sampling}}$ \\
          \hline
          $100000$ & $5000$ & $6$ & $50$ \\
          \hline
      \end{tabular}
      \caption{The parameters determining the data volume of this experiment. where $N_p$  is the number of particles, $N_{\textrm{ts}}$ is the number of time-steps saved, $N_{\textrm{phase}}$ is the number of phase space coordinates, and $N_{\textrm{sampling}}$ is the number of Monte-Carlo samplings of the initial conditions.}
      \label{tab:data_volume}
    \end{table}
